The Role of Ice Compositions and Morphology for Snowlines and the C/N/O Ratios in Active Disks

### 2016-03-03 tex/draft2/ms_edited.tex
update

\begin{abstract}
The elemental compositions of planets define their chemistry, and could potentially be used as beacons for their formation location if the elemental gas and grain ratios of planet birth environments, i.e. protoplanetary disks, are well understood. In disks, the ratios of volatile elements, such as C/O and N/O, are regulated by the abundance of the main C, N, O carriers, the ice environment in which these carriers reside, and the presence of snowlines of major volatiles at different distances from the central star. We explore the effects of dynamical processes, molecular compositions and abundances, and the ice morphology of dust grains in disks on the snowline locations of the main C, O and N carriers, and their consequences for the C/N/O ratios in gas and dust throughout the disk. 
%We find that radial drift and accretion alone can reduce the snowline radii of the main C, O and N carriers, i.e. H$_2$O, CO$_2$, CO and N$_2$, by 40-60\% compared to static disks. 
We find that if CO and N$_2$ are water dominated instead of pure ices, their snowlines move inward by $\sim$$70$\%. Both of these effects substantially change the disk regions where gas phase C/O and N/O are enhanced over the stellar value, and taken together they may move the CO and N$_2$ snowlines by a factor of $\sim$7 inward. Between the CO$_2$ and CO snowlines, the gaseous C/O and N/O are enhanced by factors of $\sim$$2$ and $\sim$$3$, respectively. The gas-phase N/O ratio is enhanced by many orders of magnitude between the CO and N$_2$ snowlines due to the depletion of oxygen gas in this region. Our estimates for the C/N/O ratios are only modestly affected by the presence of some C in the form of CH$_4$ and of some N in the form of NH$_3$.

@@

 %However, both observations and chemical models (refs) have shown that other volatiles are abundant in disks, for example nitrogen bearing species and hydrocarbons, thus shaping the disk composition.

Moreover, snowline locations strongly depend on the ice grain morphology. Laboratory experiments (\citealt{collings03}, \citealt{oberg05}, \citealt{bisschop06}, \citealt{fayolle16}) have shown that volatiles such as CO and N$_2$ have significantly different binding energies depending on whether they are pure or water dominated ices. This implies that ices in different environments will sublimate at different radii, which will substantially change the disk regions where these volatiles are present in gaseous or solid form (see Sections \ref{sec:CH4} and \ref{sec:N}).

 It follows that there are several important snowlines in disks, determined both by the disk composition and ice morphology. Figure \ref{fig:CNOstatic} shows an example of the H$_2$O, CO$_2$, CO, CH$_4$, NH$_3$ and N$_2$ snowlines in a static disk, with CO and N$_2$ as pure and water dominated ices, in the top and bottom panel, respectively. The ordinate displays the total carbon, oxygen and nitrogen abundance in solids as a function of the hydrogen total abundance. We note that we display this Figure here for qualitative purposes; the quantitative choices that we make for volatile abundances are discussed in Sections \ref{sec:CH4} and \ref{sec:N}. As expected, the total grain abundance increases with semimajor axis, as more and more species sublimate. More importantly, the CO and N$_2$ snowlines move several tens of AU inward if the ices are water dominated rather than pure. This changes the chemical abundances both in gas and dust throughout the disk, directly affecting the compositions of nascent giant planets forming in situ.

@@

This paper is organized as follows. In Section \ref{sec:review}, we review the drift-desorption model developed in Paper I. We discuss the effect of different abundances of the main carbon and oxygen carriers, grain morphology and dynamical processes on the snowline locations of carbon and oxygen bearing species and the C/O ratio in Section \ref{sec:CH4}. In Section \ref{sec:N}, we perform the same analysis for the main nitrogen carriers and the nitrogen-to-oxygen (N/O) ratio. We summarize our findings in Section \ref{sec:summary}.

@@

\section{The Effect of CO Ice Morphology and CH$_4$ on the C/O Ratio}
\label{sec:CH4}

\subsection{Role of CH$_4$}

Both in Solar system comets and in protoplanetary disks, carbon and oxygen are primarily contained in H$_2$O, CO$_2$ and CO (e.g., \citealt{rodgers02}, \citealt{lodders03}, \citealt{pontoppidan06}). However, some fraction of the carbon abundance may also be carried by CH$_4$ (e.g., \citealt{mumma96}), which may change the C/O ratio in gas and in dust throughout the disk. To quantify the magnitude of this effect, we use measured CH$_4$ abundances in protostellar cores from the \textit{Spitzer} c2d Legacy ice survey \citep{evans03}. We explore the parameter space of possible CH$_4$ abundances by assuming three different scenarios: (1) no CH$_4$, (2) the median CH$_4$ observed abundance (hereafter CH$_4$-mid), and (3) the maximum CH$_4$ observed abundance (hereafter CH$_4$-max). Thus $n_{\rm CH_4-mid}=0.0555 \times n_{\rm H_2O}$ \citep{oberg11a} and $n_{\rm CH_4-max}=0.13 \times n_{\rm H_2O}$ \citep{oberg08}, where $n_{\rm H_2O}$ is the total H$_2$O abundance. Similarly to Paper I, we use the H$_2$O, CO$_2$ and CO abundances of \citet{oberg11}. Since the abundance of carbon grains is uncertain, we assume that all the carbon that is not in the form of CH$_4$, CO and CO$_2$ is in carbon grains, so that we reproduce the Solar C/O ratio (gas+dust) of 0.54.

We determine the location of the H$_2$O, CO$_2$, CO and CH$_4$ snowlines in our static disk by balancing desorption with readsorption, following \citet{hollenbach09}. The binding energies of H$_2$O, CO$_2$, CO and CH$_4$ as pure ices are 5800 K, 2000 K, 834 K and 1300 K, respectively (\citealt{fraser01}, \citealt{collings04}, \citealt{fayolle16}, \citealt{garrod06}).

@@

Figure \ref{fig:COstatic} shows the C/O ratio in gas and dust as a function of semimajor axis in a static disk, for different CH$_4$ abundances as outlined above. As in \citet{oberg11} and Paper I, a gaseous C/O ratio of unity can be achieved beyond the CO$_2$ snowline, where oxygen gas is significantly depleted (top panel). The gas-phase C/O ratio may be further enhanced between the CO$_2$ and CH$_4$ snowlines due to the presence of additional carbon gas from CH$_4$. In this region, the C/O ratio increases by 3\% for CH$_4$-mid and by 8\% for  CH$_4$-max, as displayed in the middle and bottom panel of Figure \ref{fig:COstatic}. Based on current observations of CH$_4$ abundances, its presence in the disk only modestly affects the C/O ratio. Given the larger uncertainties in overall volatile abundances, we can neglect CH$_4$ when estimating the C/O ratio in disks. %\footnote{This may not be the case, however, in a viscous disk, as we show below.}.

\subsection{CO Ice Environment and Disk Dynamics}

As noted in Section \ref{sec:intro}, the CO binding energy varies significantly depending on the morphology of the icy grains. If CO ice is in a water dominated environment, its binding energy will be larger than in the pure ice case (834 K) due to the higher binding energy of CO to H$_2$O than to other CO molecules. \citet{fayolle16} find a CO binding energy of 1388 K in the water dominated ice scenario. We use the model of Section \ref{sec:review} to estimate the movement of the CO snowline for different grain morphologies in a viscous disk.

Figure \ref{fig:CO_ratio} shows the H$_2$O, CO$_2$ and CO snowline locations for particles with initial sizes $\sim0.06$ cm $\lesssim s \lesssim$ 7 m as well as estimates for the C/O ratio in gas and dust in a viscous disk, with the CO snowline calculated under different grain morphologies as noted above. The true snowline for particles that desorb outside the static snowline is the static snowline itself, hence desorbing particles with $s<0.06$ cm do not form true snowlines. As calculated in Paper I, drift and gas accretion may move the snowlines inwards by up to 40-60\% compared to a static disk, and specifically by up to $\sim50$ \% in the case of the CO snowline. This result is preserved for the updated CO binding energies, both for pure and water dominated ices. However, a water dominated environment moves the CO snowline inwards significantly: for our fiducial disk model, $r_{\rm CO, water} \approx 8.7$ AU. Thus if CO is water dominated instead of pure ice, the CO snowline may move inward by up to 70\%. By taking into account both disk dynamics and of CO in a water dominated environment, we find that the CO snowline may move inward by a factor of $\sim$$7$ compared to a static disk with CO as pure ice.  This large inward movement of the CO snowline implies that C/O ratios of order unity may be reached much closer to the host star if CO is a water dominated ice, and the CO snowline may be inside 10 AU for certain disk parameters.

%\emgr{Discuss observed abundances for CH4 and the choices that we make (no CH4, median value, maximum value). State that desorption energies for H2O, CO2 and CH4 are well constrained experimentally, and that the CO2 and CH4 binding energies are only weakly dependent on whether it's pure CO2/CH4 or combined with H2O, but that is not the case for CO (and N2 as we will show in the next section). Present new binding energies for CO as pure ice and mixed with water. Show Figure 1 and discuss how different CH4 abundances and binding energies affect snowline locations and C/O ratio: CO-H2O mixture (though I think it's rather CO layered on top of H2O) moves the CO snowline inward by $\sim$40 AU (will calculate percentages too); the maximum reasonable abundance of CH4 changes the C/O ratio by less than 10\%. Show Figure 2 and quantify the effect of drift and accretion on the CH4 snowline compared to a static disk. While CH4 has only a modest effect on the C/O ratio in a static disk, this effect may be larger in a viscous disk, as the C gas abundances inside the CH4 snowline may be enhanced due to the differential motion of the desorbed ices and overall nebular gas (refer to Paper I). In this study, however, we neglect these effects and therefore do not include CH4 in estimating the C/O ratio (as an aside, the figures that include CH4 in the C/O ratio with drift and desorption are quite messy due to snowlines overlapping). Show Figure 3 and estimate the difference between CO-H2O and CO pure ice snowlines in the case of drift and accretion, as well as the comparisons for the static disk for the CO snowline.}

@@

\section{The Effect of N$_2$ Ice Morphology on the N/O Ratio}
\label{sec:N}

\subsection{Nitrogen Carriers and the Role of NH$_3$}

In addition to carbon and oxygen, nitrogen is another abundant volatile in the Solar system and in disks. Chemical models of the protostellar nebula (e.g., \citealt{owen01}) and of protoplanetary disks (e.g., \citealt{rodgers02}) suggest that N$_2$ was the dominant form of nitrogen, and that giant planets have accreted their nitrogen content primarily as N$_2$ \citep{mousis14}. Observations of Solar system bodies such as Titan and Pluto show that N$_2$ is prevalent in their atmospheres (\citealt{cruikshank93}, \citealt{owen93}).  Moreover, the Rosetta spacecraft has recently made the first direct measurement of the N$_2$ abundance in comet 67P/Churyumov-Gerasimenko \citep{rubin15}. In addition to N$_2$, a fraction of the nitrogen abundances may be also carried by NH$_3$ (\citealt{bottinelli10}, \citealt{mumma11}).

Because of the high volatility of N$_2$, the gas phase nitrogen-to-oxygen (N/O) ratio in the outer disk
may be even more enhanced than the C/O ratio compared to its average value in the disk. Giant planets that form at wide separations should thus have an excess of nitrogen in their atmospheres, which could be used to trace their formation origin.  In this study, we quantify this effect in protoplanetary disks. We assume that the main nitrogen-bearing species are N$_2$ and NH$_3$, since other volatiles that contain nitrogen have significantly lower abundances in comparison (e.g., \citealt{mumma11}). We use the measured total nitrogen abundance in the Solar system, $n_{\rm N}=8 \times 10^{-5} n_{\rm H}$ \citep{lodders03}, where $n_{\rm H}$ is the hydrogen abundance in the disk midplane. Similarly to the case of CH$_4$ (see Section \ref{sec:CH4}), we explore the parameter space of possible NH$_3$ abundances using data from the Spitzer c2d Legacy ice survey, as follows: (1) no NH$_3$, (2) the median NH$_3$ observed abundance $n_{\rm NH_3-mid}=0.055 \times n_{\rm H_2O}$ \citep{oberg11a}, and (3) the maximum observed NH$_3$ abundance $n_{\rm NH_3-max}=0.1537 \times n_{\rm H_2O}$ \citep{bottinelli10}. In each case, the N$_2$ abundance then simply follows as $n_{\rm N_2}=(n_{\rm N}-n_{\rm NH_3})/2$. We determine the locations of the N$_2$ and NH$_3$ snowlines by balancing desorption with readsorption \citep{hollenbach09}, with N$_2$ and NH$_3$ pure ice binding energies of  767 K and  2965 K, respectively (\citealt{fayolle16}, \citealt{martin14}).

Figure \ref{fig:Nstatic} shows the snowline locations of the main oxygen and nitrogen carriers and the N/O ratio in gas and dust as a function of semimajor axis in a static disk, for our three choices of the NH$_3$ abundance. For comparison, the horizontal dashed line shows the average N/O ratio in the disk. As expected, the gaseous N/O ratio generally exhibits an increasing trend towards the outer disk as more oxygen gas is depleted, with small decreases between the NH$_3$ and CO$_2$ snowlines (by 16\% for NH$_3$-mid and by 18\% for NH$_3$-max, respectively) due to NH$_3$ freeze-out. More importantly, the gas-phase N/O ratio in the outer disk is enhanced by more than a factor of three compared to its average value. This enhancement is more pronounced than the C/O gas-phase enhancement of a factor of two in the outer disk (see Figure \ref{fig:COstatic}). The N/O ratio reaches particularly high values between the CO and N$_2$ snowlines, where all the oxygen is now contained in grains.

@@

%\emgr{Similar to the previous section, but with more details. Discuss that nitrogen is abundant in the solar system and disks and primarily found as N2. Due to the high volatility of N2, the gas phase N/O ratio in the outer disk may be even more enhanced than the C/O ratio. A fraction of the nitrogen abundance may be also carried by NH3. Discuss NH3 observed abundances and the choices that we make (no NH3, median, maximum). State that the NH3 desorption energy is only weakly dependent on whether it's pure NH3 or combined with H2O, but that is not the case for N2. Present new binding energies for N2 as pure ice and combined with water. Show Figure 4 and discuss how different nitrogen abundances and binding energies affect snowline locations and N/O ratio: N2 combined with H2O moves the N2 snowline inward by $\sim$50 AU (will calculate percentages too); the maximum reasonable abundance of NH3 changes the N/O ratio by $\sim$15\%. In the outer disk, the N/O ratio is enhanced by a factor of $\sim$4 compared to the solar value, twice as much as the C/O enhancement. Show Figure 5 and quantify the effect of drift and accretion on the NH3 snowline compared to a static disk. While NH3 does not have a significant effect on the N/O ratio in a static disk, this effect may be larger in a viscous disk, as the N gas abundance inside the NH3 snowline may be enhanced due to the differential motion of the desorbed ices and overall nebular gas (refer to Paper I). In this study, however, we neglect these effects and therefore do not include NH3 in estimating the N/O ratio (again, N/O ratio figure with drift is quite messy when including NH3 and does not add any information that is not already shown in Figures 4 and 5). Show Figure 6 and estimate the difference between N2-H2O and N2 pure ice snowlines in the case of drift, as well as the comparisons for the static disk for the N2 snowline. State that there will be an overabundance of gas-phase N/O between the CO and N2 snowlines, as there is no oxygen gas in this region.}

%\begin{figure}[h!]
%\centering
%\includegraphics[width=0.5\textwidth]{../../figs/desorption_distance_NH3_steady.pdf}
%%\vspace{-0.5in}
%\caption{Desorption distance as a function of initial distance for NH3.... Drift and gas accretion move the NH3 snowline inward by x\%.} 
%\label{fig:NH3}
%\end{figure}

\subsection{N$_2$ Ice Environment and Disk Dynamics}

>>>>

\begin{abstract}
The elemental compositions of planets define their chemistry, and could potentially be used as beacons for their formation location if the elemental gas and grain ratios of planet birth environments, i.e. protoplanetary disks, are well understood. In disks, the ratios of volatile elements, such as C/O and N/O, are regulated by the abundance of the main C, N, O carriers, their ice binding environment, and the presence of snowlines of major volatiles at different distances from the central star. We explore the effects of dynamical processes, molecular compositions and abundances, and the ice morphology of dust grains in disks on the snowline locations of the main C, O and N carriers, and their consequences for the C/N/O ratios in gas and dust throughout the disk. 
%We find that radial drift and accretion alone can reduce the snowline radii of the main C, O and N carriers, i.e. H$_2$O, CO$_2$, CO and N$_2$, by 40-60\% compared to static disks. 
We find that if CO and N$_2$ are water dominated instead of pure ices, their snowlines move inward by $\sim$$70$\%. Both ice morphology and disk dynamics substantially change the disk regions where gas phase C/O and N/O are enhanced over the stellar value, and taken together they may move the CO and N$_2$ snowlines by a factor of $\sim$7 inward. Between the CO$_2$ and CO snowlines, the gaseous C/O and N/O are enhanced by factors of $\sim$$2$ and $\sim$$3$, respectively. The gas-phase N/O ratio is enhanced by many orders of magnitude between the CO and N$_2$ snowlines due to the depletion of oxygen gas in this region. Our estimates for the C/N/O ratios are moderately affected by the presence of some C in the form of CH$_4$ and of some N in the form of NH$_3$.

@@

Both in Solar system comets and in protoplanetary disks carbon and oxygen are primarily contained in H$_2$O, CO$_2$ and CO (e.g., \citealt{rodgers02}, \citealt{lodders03}, \citealt{pontoppidan06}). However, some fraction of the carbon abundance may also be carried by CH$_4$ (e.g., \citealt{mumma96}), which may change the C/O ratio in gas and in dust throughout the disk. Nitrogen is another abundant volatile in the Solar system and in disks. Chemical models of the protostellar nebula (e.g., \citealt{owen01}) and of protoplanetary disks (e.g., \citealt{rodgers02}) suggest that N$_2$ was the dominant form of nitrogen, and that giant planets have accreted their nitrogen content primarily as N$_2$ \citep{mousis14}. Observations of Solar system bodies such as Titan and Pluto show that N$_2$ is prevalent in their atmospheres (\citealt{cruikshank93}, \citealt{owen93}).  Moreover, the Rosetta spacecraft has recently made the first direct measurement of the N$_2$ abundance in comet 67P/Churyumov-Gerasimenko \citep{rubin15}. In addition to N$_2$, a fraction of the nitrogen abundances may be also carried by NH$_3$ (\citealt{bottinelli10}, \citealt{mumma11}). Because of the high volatility of N$_2$, the gas phase nitrogen-to-oxygen (N/O) ratio in the outer disk may be even more enhanced than the C/O ratio compared to its average value in the disk. Giant planets that form at wide separations should thus have an excess of nitrogen in their atmospheres, which could be used to trace their formation origin.

Moreover, snowline locations strongly depend on the ice grain morphology. Laboratory experiments (\citealt{collings03}, \citealt{oberg05}, \citealt{bisschop06}, \citealt{fayolle16}) have shown that volatiles such as CO and N$_2$ have significantly different binding energies depending on whether they are pure or water dominated ices. This implies that ices in different environments will sublimate at different radii, which will substantially change the disk regions where these volatiles are present in gaseous or solid form (see Section \ref{sec:static}).

%It follows that there are several important snowlines in disks, determined both by the disk composition and ice morphology. Figure \ref{fig:CNOstatic} shows an example of the H$_2$O, CO$_2$, CO, CH$_4$, NH$_3$ and N$_2$ snowlines in a static disk, with CO and N$_2$ as pure and water dominated ices, in the top and bottom panel, respectively. The ordinate displays the total carbon, oxygen and nitrogen abundance in solids as a function of the hydrogen total abundance. We note that we display this Figure here for qualitative purposes; the quantitative choices that we make for volatile abundances are discussed in Sections \ref{sec:CH4} and \ref{sec:N}. As expected, the total grain abundance increases with semimajor axis, as more and more species sublimate. More importantly, the CO and N$_2$ snowlines move several tens of AU inward if the ices are water dominated rather than pure. This changes the chemical abundances both in gas and dust throughout the disk, directly affecting the compositions of nascent giant planets forming in situ.

@@

This paper is organized as follows. In Section \ref{sec:review}, we review the drift-desorption model developed in Paper I. We discuss the effect of different abundances of the main carbon, oxygen and nitrogen carriers, grain morphology and disk dynamics on snowline locations and the C/N/O ratios in Section \ref{sec:results}. We discuss the implications of our results in Section \ref{sec:discussion} and summarize our findings in Section \ref{sec:summary}.

@@

\section{Results}
\label{sec:results}

\subsection{Snowlines in a Static Disk: The Importance of Ice Morphology}
\label{sec:snowlines}

As we note in Section \ref{sec:intro}, the disk volatile composition and the ice morphology determine the location of several important snowlines. In this work we focus on the primary carbon, oxygen and nitrogen carriers, i.e. H$_2$O, CO$_2$, CO, N$_2$, and to a lesser extent, CH$_4$ and NH$_3$. To qualify the effect of the presence of some carbon in the form of CH$_4$, we use measured CH$_4$ abundances in protostellar cores from the \textit{Spitzer} c2d Legacy ice survey \citep{evans03}. We explore the parameter space of possible CH$_4$ abundances by assuming three different scenarios: (1) no CH$_4$, (2) the median CH$_4$ observed abundance (hereafter CH$_4$-mid), and (3) the maximum CH$_4$ observed abundance (hereafter CH$_4$-max). Thus $n_{\rm CH_4-mid}=0.0555 \times n_{\rm H_2O}$ \citep{oberg11a} and $n_{\rm CH_4-max}=0.13 \times n_{\rm H_2O}$ \citep{oberg08}, where $n_{\rm H_2O}$ is the total H$_2$O abundance. Similarly to Paper I, we use the H$_2$O, CO$_2$ and CO abundances of \citet{oberg11}. Since the abundance of carbon grains is uncertain, we assume that all the carbon that is not in the form of CH$_4$, CO and CO$_2$ is in carbon grains, so that we reproduce the Solar C/O ratio (gas+dust) of 0.54.

We assume that the main nitrogen-bearing species are N$_2$ and NH$_3$, since other volatiles that contain nitrogen have significantly lower abundances in comparison (e.g., \citealt{mumma11}). We use the measured total nitrogen abundance in the Solar system, $n_{\rm N}=8 \times 10^{-5} n_{\rm H}$ \citep{lodders03}, where $n_{\rm H}$ is the hydrogen abundance in the disk midplane. Similarly to the case of CH$_4$, we explore the parameter space of possible NH$_3$ abundances using data from the Spitzer c2d Legacy ice survey, as follows: (1) no NH$_3$, (2) the median NH$_3$ observed abundance $n_{\rm NH_3-mid}=0.055 \times n_{\rm H_2O}$ \citep{oberg11a}, and (3) the maximum observed NH$_3$ abundance $n_{\rm NH_3-max}=0.1537 \times n_{\rm H_2O}$ \citep{bottinelli10}. In each case, the N$_2$ abundance then simply follows as $n_{\rm N_2}=(n_{\rm N}-n_{\rm NH_3})/2$.

We determine the location of the H$_2$O, CO$_2$, CO, CH$_4$, N$_2$ and NH$_3$ snowlines in our static disk by balancing desorption with readsorption, following \citet{hollenbach09}. The binding energies of H$_2$O, CO$_2$, CO, CH$_4$, N$_2$ and NH$_3$ as pure ices are 5800 K, 2000 K, 834 K, 1300 K, 767 K and 2965 K, respectively (\citealt{fraser01}, \citealt{collings04}, \citealt{fayolle16}, \citealt{garrod06}, \citealt{martin14}). For CO and N$_2$ as water dominated ices the binding energies are 1388 K and 1266 K, respectively \citep{fayolle16}. Figure \ref{fig:CNOstatic} shows the resulting snowline locations, with CO and N$_2$ pure ices (top panel) and water dominated ices (bottom panel). The ordinate displays the total carbon, oxygen and nitrogen abundance in solids as a function of the hydrogen total abundance.  We assume that the CH$_4$ and NH$_3$ abundances are $n_{\rm CH_4-mid}$ and $n_{\rm NH_3-mid}$.  As expected, the total grain abundance increases with semimajor axis, as more and more species sublimate. More importantly, the CO and N$_2$ snowlines move several tens of AU inward if the ices are water dominated rather than pure. This changes the chemical abundances both in gas and dust throughout the disk, directly affecting the compositions of nascent giant planets forming in situ.

%It follows that there are several important snowlines in disks, determined both by the disk composition and ice morphology. Figure \ref{fig:CNOstatic} shows an example of the H$_2$O, CO$_2$, CO, CH$_4$, NH$_3$ and N$_2$ snowlines in a static disk, with CO and N$_2$ as pure and water dominated ices, in the top and bottom panel, respectively.

\subsection{C/N/O Ratios in Static Disks}
\label{sec:static}

In this section we determine the C/O and N/O ratios in gas and dust throughout our static disk, and the extent by which they are affected by the presence of CH$_4$ and NH$_3$. Since here we are primarily interested in the magnitude of the C/N/O ratios rather than snowline locations (which have been discussed extensively in Section \ref{sec:snowlines}), we assume for simplicity that CO and N$_2$ are pure ices.

%Both in Solar system comets and in protoplanetary disks, carbon and oxygen are primarily contained in H$_2$O, CO$_2$ and CO (e.g., \citealt{rodgers02}, \citealt{lodders03}, \citealt{pontoppidan06}). However, some fraction of the carbon abundance may also be carried by CH$_4$ (e.g., \citealt{mumma96}), which may change the C/O ratio in gas and in dust throughout the disk. 
%To quantify the magnitude of this effect, we use measured CH$_4$ abundances in protostellar cores from the \textit{Spitzer} c2d Legacy ice survey \citep{evans03}. We explore the parameter space of possible CH$_4$ abundances by assuming three different scenarios: (1) no CH$_4$, (2) the median CH$_4$ observed abundance (hereafter CH$_4$-mid), and (3) the maximum CH$_4$ observed abundance (hereafter CH$_4$-max). Thus $n_{\rm CH_4-mid}=0.0555 \times n_{\rm H_2O}$ \citep{oberg11a} and $n_{\rm CH_4-max}=0.13 \times n_{\rm H_2O}$ \citep{oberg08}, where $n_{\rm H_2O}$ is the total H$_2$O abundance. Similarly to Paper I, we use the H$_2$O, CO$_2$ and CO abundances of \citet{oberg11}. Since the abundance of carbon grains is uncertain, we assume that all the carbon that is not in the form of CH$_4$, CO and CO$_2$ is in carbon grains, so that we reproduce the Solar C/O ratio (gas+dust) of 0.54.

%We determine the location of the H$_2$O, CO$_2$, CO, CH$_4$, N$_2$ and NH$_3$ snowlines in our static disk by balancing desorption with readsorption, following \citet{hollenbach09}. The binding energies of H$_2$O, CO$_2$, CO, CH$_4$ and N$_2$ as pure ices are 5800 K, 2000 K, 834 K, 1300 K, respectively (\citealt{fraser01}, \citealt{collings04}, \citealt{fayolle16}, \citealt{garrod06}).

@@

Figure \ref{fig:COstatic} shows the C/O ratio in gas and dust as a function of semimajor axis in a static disk, for three different CH$_4$ abundances as outlined above: no CH$_4$ (top panel), CH$_4$-mid (middle panel) and CH$_4$-max (bottom panel). As in \citet{oberg11} and Paper I, a gaseous C/O ratio of unity can be achieved between the CO$_2$ and CO snowlines, where oxygen gas is significantly depleted. The gas-phase C/O ratio may be further enhanced between the CO$_2$ and CH$_4$ snowlines due to the presence of additional carbon gas from CH$_4$. In this region, the C/O ratio increases by 3\% for CH$_4$-mid and by 8\% for CH$_4$-max, as displayed in the middle and bottom panels of Figure \ref{fig:COstatic}. Based on the range of observed CH$_4$ protostellar abundances, its presence in the disk only modestly affects the C/O ratio. Given the larger uncertainties in overall volatile abundances, we can neglect CH$_4$ when estimating the C/O ratio in disks.

%Nitrogen is another abundant volatile in the Solar system and in disks. Chemical models of the protostellar nebula (e.g., \citealt{owen01}) and of protoplanetary disks (e.g., \citealt{rodgers02}) suggest that N$_2$ was the dominant form of nitrogen, and that giant planets have accreted their nitrogen content primarily as N$_2$ \citep{mousis14}. Observations of Solar system bodies such as Titan and Pluto show that N$_2$ is prevalent in their atmospheres (\citealt{cruikshank93}, \citealt{owen93}).  Moreover, the Rosetta spacecraft has recently made the first direct measurement of the N$_2$ abundance in comet 67P/Churyumov-Gerasimenko \citep{rubin15}. In addition to N$_2$, a fraction of the nitrogen abundances may be also carried by NH$_3$ (\citealt{bottinelli10}, \citealt{mumma11}).

%Because of the high volatility of N$_2$, the gas phase nitrogen-to-oxygen (N/O) ratio in the outer disk
%may be even more enhanced than the C/O ratio compared to its average value in the disk. Giant planets that form at wide separations should thus have an excess of nitrogen in their atmospheres, which could be used to trace their formation origin.  
%In this study, we quantify this effect in protoplanetary disks. We assume that the main nitrogen-bearing species are N$_2$ and NH$_3$, since other volatiles that contain nitrogen have significantly lower abundances in comparison (e.g., \citealt{mumma11}). We use the measured total nitrogen abundance in the Solar system, $n_{\rm N}=8 \times 10^{-5} n_{\rm H}$ \citep{lodders03}, where $n_{\rm H}$ is the hydrogen abundance in the disk midplane. Similarly to the case of CH$_4$ (see Section \ref{sec:CH4}), we explore the parameter space of possible NH$_3$ abundances using data from the Spitzer c2d Legacy ice survey, as follows: (1) no NH$_3$, (2) the median NH$_3$ observed abundance $n_{\rm NH_3-mid}=0.055 \times n_{\rm H_2O}$ \citep{oberg11a}, and (3) the maximum observed NH$_3$ abundance $n_{\rm NH_3-max}=0.1537 \times n_{\rm H_2O}$ \citep{bottinelli10}. In each case, the N$_2$ abundance then simply follows as $n_{\rm N_2}=(n_{\rm N}-n_{\rm NH_3})/2$. We determine the locations of the N$_2$ and NH$_3$ snowlines by balancing desorption with readsorption \citep{hollenbach09}, with N$_2$ and NH$_3$ pure ice binding energies of  767 K and  2965 K, respectively (\citealt{fayolle16}, \citealt{martin14}).

Figure \ref{fig:Nstatic} shows the snowline locations of the main oxygen and nitrogen carriers and the N/O ratio in gas and dust as a function of semimajor axis in a static disk, for our three choices of the NH$_3$ abundance (see Section \ref{sec:snowlines}): no NH$_3$ (top panel), NH$_3$-mid (middle panel) and NH$_3$-max (bottom panel). For comparison, the horizontal dashed lines show the average N/O ratio in the disk. As expected, the gaseous N/O ratio generally exhibits an increasing trend towards the outer disk as more oxygen gas is depleted, with small decreases between the NH$_3$ and CO$_2$ snowlines (by 16\% for NH$_3$-mid and by 18\% for NH$_3$-max, respectively) due to NH$_3$ freeze-out. While the presence of NH$_3$ only moderately affects our results for the N/O ratio, NH$_3$ is important since otherwise there would be no solid nitrogen inside the N$_2$ snowline, which would be contrary to asteroid and comet observations (\citealt{wyckoff91}, \citealt{mumma11}, \citealt{bergin15}).

The gas-phase N/O ratio is enhanced by a factor of two outside the H$_2$O snowline compared to its average value, by more than a factor of three between the CO$_2$ and CO snowlines, and by orders of magnitude between the CO and N$_2$ snowlines, which can span tens of AU depending on disk parameters and the relative CO and N$_2$ ice binding environment. This enhancement is more pronounced than the C/O gas-phase enhancement of a factor of two in the outer disk (see Figure \ref{fig:COstatic}). %The N/O ratio reaches particularly high values between the CO and N$_2$ snowlines, where all the oxygen is now contained in grains.

\subsection{C/N/O Ratios in Dynamic Disks}
\label{sec:dynamic}

%As noted in Section \ref{sec:intro}, the CO binding energy varies significantly depending on the morphology of the icy grains. If CO ice is in a water dominated environment, its binding energy will be larger than in the pure ice case (834 K) due to the higher binding energy of CO to H$_2$O than to other CO molecules. \citet{fayolle16} find a CO binding energy of 1388 K in the water dominated ice scenario.

Here we use the model of Section \ref{sec:review} to estimate the movement of the CO and N$_2$ snowlines for different grain morphologies in a viscous disk. Figure \ref{fig:CO_ratio} shows the H$_2$O, CO$_2$ and CO snowline locations for particles with initial sizes $\sim0.06$ cm $\lesssim s \lesssim$ 7 m as well as estimates for the C/O ratio in gas and dust in a viscous disk, with the CO snowline calculated under different grain morphologies as noted above. The true snowline for particles that desorb outside the static snowline is the static snowline itself, hence desorbing particles with $s<0.06$ cm do not form true snowlines. If the CO binding environment is known, the CO snowline moves inward by up to $\sim$50 \% compared to a static disk for each case (pure and water dominated ices) due to disk dynamics. The full range of potential CO snowlines taking into account both ice morphology and disk dynamics span $\sim$8.7 AU to $\sim$61 AU, which is a factor of $\sim$7 difference. This implies that gas phase C/O ratios of order unity may be reached in the giant planet forming zone, and the CO snowline may be inside 10 AU for certain disk parameters.

@@

%As in the case for CO, the N$_2$ binding energy is also strongly dependent on the ice environment in which the grain resides. If N$_2$ is a water dominated ice, the N$_2$ binding energy is 1266 K \citep{fayolle16}.

Figure \ref{fig:NO_ratio} shows the H$_2$O, CO$_2$, CO and N$_2$ snowline locations in a viscous disk for the same range of initial particle sizes as in Figure \ref{fig:CO_ratio}, and with the CO and N$_2$ snowlines calculated assuming different grain morphologies as explained above, as well as estimates for the N/O ratio throughout the disk. For simplicity, we assume that all nitrogen is bound in N$_2$. This choice is justified since the presence of some NH$_3$ only moderately changes the N/O ratio (see Figure \ref{fig:Nstatic}), and since we are primarily interested in the N$_2$ snowline locations rather than exact values for the N/O ratio. The innermost N$_2$ snowlines in the viscous disk, created by particles with $s \sim 7$ m for our fiducial model, are located at $r_{\rm N_2, pure} \approx 42$ AU for N$_2$ as pure ice and at $r_{\rm N_2, water} \approx 11$ AU for N$_2$ in water dominated ices. Thus for each case (pure versus water dominated ices), the N$_2$ snowline moves inward by up to 50\% due to disk dynamics. By taking into account both ice morphology and disk dynamics, the full range of potential N$_2$ snowlines span $\sim$11 to $\sim$79 AU, which is a factor of $\sim$7 difference. Similarly to the case for CO, the N$_2$ snowline may be close to or inward of 10 AU for certain disk models.

@@
